A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-07-10 paper/theor-experiment.tex
adding final data to paper

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig1) at (0,0)
	{\includegraphics[width=0.9\textwidth]{Report/awa-drawing}};
	\node[fill=white, inner sep=2pt] (txt2) at (35,15) {Interferometer};
	\node[fill=white, inner sep=2pt, rotate=26] (txt2) at (18,19.5) {Slit};	
	\node[fill=white, inner sep=2pt, rotate=20] (txt2) at (13,27) {Window};
	\end{tikzpicture}
	\caption{High charge beam line layout at the AWA.}
	\label{awa-pic}
\end{figure}
%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, 
along with three constraints.
The objectives include transverse and longitudinal beam size, 
transverse momentum and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_1=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-pic}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

@@

\subsubsection{Preliminary Tests} \label{awa:subsection:test}
\nrnote{What else could we name this subsection?}
A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size $16 \times 16 \times 32$ was chosen, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. The
average run time of each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes.
%A summary of the parameters used is listed in Table \ref{fidelity}.

@@

While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals, 
and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem},
\nrnote{Changed variables for final run, the variables in 5.3 no longer match these small runs. Need to rework this section} 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

@@

After collection of the data for all four experiments, several metrics
were compared, including number of generations completed in twenty four hours,
Pareto fronts at $s_1$ and $s_2$, and the design variables at the Pareto front. 
From Table \ref{extable}, we clearly see ex-3 is significantly 
slower, as it evaluated about 30 to 40 fewer generations than the other experiments.
Perhaps this trade off would be acceptable if the Pareto front was significantly 
improved, but from Fig. \ref{expareto}, we can see this is not the case.
The Pareto fronts at $s_1$, are nearly identical. We could expect
they would continue to converge given more time. 
When looking at the Pareto front for $s_2$, we see ex-4 has a slight 
edge over the others.  
\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{Report/ex-pareto1}%
	\llap{\raisebox{1.8cm}{%  move next graphics to top right corner
			\includegraphics[height=5.5cm]{Report/ex-pareto2}%
	}}
	\caption{My overlay figure.}
\end{figure}

Next compare the design variables at each point on the 
Pareto front.

@@

The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers were in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources. 
This would be the population sizes used for all subsequent runs.

\subsubsection{AWA Optimization Results}
With the simulation, GA, and node resource parameters set by the work in \ref{awa:subsection:test}, 
the full optimization problem described in \ref{awaproblem} was run for 200 generations.

>>>>

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig1) at (0,0)
	{\includegraphics[width=0.9\textwidth]{Report/awa-drawing}};
	\node[fill=white, inner sep=2pt] (txt2) at (35,15) {Interferometer};
	\node[fill=white, inner sep=2pt, rotate=26] (txt2) at (18,19.5) {Slit};	
	\node[fill=white, inner sep=2pt, rotate=20] (txt2) at (13,27) {Window};
	\end{tikzpicture}
	\caption{High charge beam line layout at the AWA.}
	\label{awa-pic}
\end{figure}
%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, 
along with three constraints.
The objectives include transverse and longitudinal beam size, 
transverse momentum and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_1=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-pic}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.
\nrnote{Maybe eqs 5.10-5.21 and code example should move after time step scan and hyper parameters}

@@

\subsubsection{Time Step Scan} \label{awa:subsection:test}
A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size $16 \times 16 \times 32$ was chosen, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. The
average run time of each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes.
%A summary of the parameters used is listed in Table \ref{fidelity}.

@@

\subsubsection{Hyper parameter Scan}
While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals, 
and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem},
\nrnote{Changed variables for final run, the variables in 5.3 no longer match these small runs. Need to rework this section} 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

@@

After collection of the data for all four experiments, several metrics
were compared, including number of generations completed in twenty four hours,
Pareto fronts at $s_1$ and $s_2$, and the design variables at the Pareto front. 
From Table \ref{extable}, we clearly see ex-3 is significantly 
slower, as it evaluated about 30 to 40 fewer generations than the other experiments.
Perhaps this trade off would be acceptable if the Pareto front was significantly 
improved, but from Fig. \ref{expareto}, we can see this is not the case.
The Pareto fronts at $s_1$, are nearly identical. We could expect
they would continue to converge given more time. 
Similar arguments can be made for ex-2, which evaluated about 15 less generations
with no clear benefit in either Pareto front.
When looking at the Pareto front at $s_2$, only ex-4 has a slight 
edge over the others.
With ex-2 and ex-3 eliminated due to evaluation time, 
the hyper parameters in ex-4 were chosen as the default values for subsequent runs.
\iffalse  
\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{Report/ex-pareto1}%
	\llap{\raisebox{1.8cm}{%  move next graphics to top right corner
			\includegraphics[height=5.5cm]{Report/ex-pareto2}%
	}}
	\caption{My overlay figure.}
\end{figure}  
\fi

@@

\subsubsection{AWA Optimization Results}
With the time steps and hyper parameters set by the work in \ref{awa:subsection:test}, 
the full optimization problem described in \ref{awaproblem} was run for 200 generations.
The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers were in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources.

Again, the location of optimization is $s_3=19.4$ [m]. 
This is the entrance to the fifth quad in the beam line. 
This location is where the beam should be captured and focused through subsequent elements.
After collection of the data, Pareto fronts comparing the objectives were plotted, 
see Figs. X-Z.
The $\sigma_x$ and $\sigma_{px}$ plots are of most interest, as they are the most heavility 
impacted by bending elements. With this in mind, several beam parameters corresponding to
options on Pareto Front in Fig. X were plotted and compared. 
The best results are shown in Fig. Z. 
The maximum beam sizes are well below the beam pip aperture limits as shown in Fig. (b).
